---
title: "DS 202 Final Project"
output: 
  github_document:
    number_section: FALSE
---

<!-- README.md is generated from README.Rmd. Please edit the README.Rmd file -->

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)


library(ggplot2)
library(dplyr)
library(readr)

# RAW DATA 
happy_score <- read.csv("Happiness_Scores.xls-Sheet1.csv")
head(happy_score)
```

This repository serves as a starter repo for your final project, and this Rmd is supposed to serve as a starter file for your project report.

# Analysis of Worldwide Ladder Score {.unnumbered}

Authors: Cassandra Larsen, Chen Sang

# Intro

The goal of this project was to find out what varibles impacted the ladder score, which is this data's indicator of a happines score. We were intrested in this topic as we are trying to find out where we wish to possibly move after the conculsion of our undergraduate degree for either a job or for a more advnced degree. We looked into the best countries to live in terms of having the happiest citizens and analyzing what factors contributed to the ladder score the most and the least, as well as looking into trends in these varibles thatg impact the laterscore

In pursuit of the stated goal, we will explore the following questions:

#### **Trend Analysis:**

-   What is the relationship between the ‘Ladder score’ and ‘GDP per capita’?

-   Does ‘Social support’ correlate strongly with the ‘Ladder score’?

-   What variable(s) have the largest impact on ladder score that are also statistically significant?

**Comparative Analysis**:

-   Which countries have the highest ‘Ladder score’?

-   How do factors like ‘Generosity’ and ‘Perceptions of corruption’ vary among the top 10 happiest countries compared to the bottom 10?

**Distribution Analysis**:

-   What is the distribution of the ‘Ladder score’ across all countries? Is it normally distributed?

-   How does the distribution of ‘Healthy life expectancy’ compare to that of ‘Freedom to make life choices’?

# Quick Data Summary

**Relevant Variables from Dataset**:

-   Ladder_score: ladder score of each country which is this data's happiness score

-   Explained_by_Log_GDP_per_capita: measure of a country's economic output that accounts for its number of people

-   Country_name: country's name

-   Explained_by_Social_support: interpersonal connections and community resources contribute to individual and collective well-being.

-   Explained_by_Generosity: measures the average prevalence of generosity within a country

-   Explained_by_Perceptions_of_corruption: level of corruption perception within political institutions and businesses

-   Explained_by_Healthy_life_expectancy: measures the average number of years that a person is expected to live in full health

-   Explained_by_Freedom_to_make_life_choices: measures the individuals' perceived freedom to make life choices, reflecting how free people feel to choose the course of their lives, including major decisions about careers, education, and family life

**Cleaning data:**

```{r}
# Remove rows with any null values
clean_Happy_data <- na.omit(happy_score)

# Replace dots with underscores in column names
names(clean_Happy_data) <- gsub("\\.+", "_", names(clean_Happy_data))

# Now the column names have underscores instead of dots
head(names(clean_Happy_data))

# View the cleaned data
head(clean_Happy_data)

```

# Results

### **Trend Analysis:**

-   **What is the relationship between the ‘Ladder score’ and ‘GDP per capita’?**

    -   A linear regression analysis can help us understand how much of the variation in **`Ladder_score`** can be explained by **`Explained_by_Log_GDP_per_capita`**

    ```{r}
    # Perform linear regression
    linearModel <- lm(Ladder_score ~ Explained_by_Log_GDP_per_capita, data = clean_Happy_data)

    # Summary of the regression model to see coefficients and statistics
    summary(linearModel)
    ```

    ```{r}
    # Plotting the data points
    plot(clean_Happy_data$Explained_by_Log_GDP_per_capita, clean_Happy_data$Ladder_score, 
         main = "Relationship between Ladder Score and Log GDP per Capita",
         xlab = "Log GDP per Capita (Explained part)",
         ylab = "Ladder Score",
         pch = 19, col = "blue")

    # Adding the regression line
    abline(linearModel, col = "red", lwd = 2)

    # Enhancing the plot with grid lines for better readability
    grid(nx = NULL, ny = NULL, col = "gray", lty = "dotted")
    ```

-   Results

    -   **Positive Relationship**: There's a clear positive correlation between Log GDP per Capita and Ladder Score. This suggests that as the economic output per capita increases, the happiness score tends to increase as well. This is indicated by the upward slope of the red regression line.

    -   **Strength of the Relationship**: The spread of the blue points around the regression line suggests that while GDP per capita explains a significant portion of the variance in happiness scores, there are also other factors that influence happiness since the points are not perfectly aligned along the line. This is consistent with the R-squared value from your regression summary (approximately 0.59), indicating that around 59% of the variation in Ladder Score can be explained by GDP per capita.

    -   **Statistical Significance**: The regression coefficients for the intercept and Log GDP per capita are both statistically significant (p-values \< 2.2e-16), indicating a very low probability that these results are due to random chance.

-   **Does ‘Social support’ correlate strongly with the ‘Ladder score’?**

    -   Pearson correlation coefficient, which quantifies the strength and direction of the relationship, and to create a scatter plot with a regression line to visually assess the relationship.

```{r}
# Calculate Pearson correlation coefficient
correlation <- cor(clean_Happy_data$Explained_by_Social_support, clean_Happy_data$Ladder_score, use = "complete.obs")
print(paste("Pearson correlation coefficient between Social Support and Ladder Score is:", correlation))
```

```{r}
# Plotting the data points
plot(clean_Happy_data$Explained_by_Social_support, clean_Happy_data$Ladder_score, 
     main = "Relationship between Social Support and Ladder Score",
     xlab = "Social Support",
     ylab = "Ladder Score",
     pch = 19, col = "blue")

# Fit a linear model and add the regression line to the plot
model <- lm(Ladder_score ~ Explained_by_Social_support, data = clean_Happy_data)
abline(model, col = "red", lwd = 2)

# Add grid lines for better readability
grid(nx = NULL, ny = NULL, col = "gray", lty = "dotted")

```

-   Results

    -   **Strength of Relationship**: A correlation coefficient of 0.81 is quite high, suggesting a strong association between the levels of social support in a country and its reported happiness (Ladder Score).

    -   **Direction of Relationship**: Since the correlation coefficient is positive, it means that as social support increases, the Ladder Score generally increases as well. This implies that countries with higher levels of social support tend to have higher happiness scores.

-   **What variable(s) have the largest impact on ladder score that are also statistically significant?**

    ```{r}
    # Fit the multiple linear regression model
    Mmodel <- lm(Ladder_score ~ Explained_by_Log_GDP_per_capita + Explained_by_Social_support + Explained_by_Healthy_life_expectancy + Explained_by_Freedom_to_make_life_choices + Explained_by_Generosity + Explained_by_Perceptions_of_corruption, data = clean_Happy_data)


    # Summary of the model to see coefficients, R-squared, etc.
    summary(Mmodel)
    ```

-   **Coefficients Analysis**

    -   **(Intercept) (1.1704)**: The expected value of **Ladder_score** when all other variables are zero, which is typically a theoretical scenario in this context.

    -   **Explained_by_Log_GDP_per_capita (0.4415, p = 0.03631)**: Indicates a positive and significant association with the ladder score. Higher GDP per capita (logged) is associated with higher happiness scores, although the p-value is close to 0.05, suggesting a weaker significance compared to other variables.

    -   **Explained_by_Social_support (1.4384, p = 3.25e-10)**: Shows a strong positive impact on happiness scores. This variable is highly significant, emphasizing the importance of social support in contributing to happiness.

    -   **Explained_by_Healthy_life_expectancy (1.3825, p = 0.00593)**: Also significantly positive, indicating that longer healthy life expectancies are closely associated with higher happiness scores.

    -   **Explained_by_Freedom_to_make_life_choices (1.8968, p = 3.27e-08)**: One of the strongest predictors, suggesting that freedom in making life choices is a crucial determinant of happiness.

    -   **Explained_by_Generosity (0.4621, p = 0.46776)**: Not a significant predictor in this model, indicating that generosity does not have a statistically significant effect on the ladder score within the scope of this dataset.

    -   **Explained_by_Perceptions_of_corruption (1.0072, p = 0.01516)**: This is significant, suggesting that perceptions of corruption negatively affect happiness scores, where less perceived corruption is associated with higher happiness.

-   As generosity was not statistically significant it was removed and we re-ran the model.

    ```{r}
    # Fit the multiple linear regression model
    Mmodel <- lm(Ladder_score ~ Explained_by_Log_GDP_per_capita + Explained_by_Social_support + Explained_by_Healthy_life_expectancy + Explained_by_Freedom_to_make_life_choices + Explained_by_Perceptions_of_corruption, data = clean_Happy_data)


    # Summary of the model to see coefficients, R-squared, etc.
    summary(Mmodel)

    ```

-   **Analysis of Coefficients**

    -   **Intercept (1.2206)**: The base level of the **Ladder_score** when all other predictors are zero. This is an abstract concept since the predictors are not likely to be zero in real scenarios.

    -   **Explained_by_Log_GDP_per_capita (Coefficient = 0.4066, p-value = 0.04701)**: This suggests that for every one-unit increase in the logarithm of GDP per capita, the **Ladder_score** increases by approximately 0.407 units, assuming other factors are held constant. The p-value is close to 0.05, making this predictor marginally significant.

    -   **Explained_by_Social_support (Coefficient = 1.4588, p-value = 1.32e-10)**: A highly significant predictor. An increase in social support is strongly associated with an increase in happiness scores, more so than GDP per capita.

    -   **Explained_by_Healthy_life_expectancy (Coefficient = 1.3939, p-value = 0.00544)**: Also a significant predictor. It indicates that longer healthy life expectancy is associated with higher happiness scores.

    -   **Explained_by_Freedom_to_make_life_choices (Coefficient = 1.9406, p-value = 9.44e-09)**: This shows a strong positive association with **Ladder_score**. It implies that greater freedom in making life choices significantly boosts happiness, even more strongly than social support.

    -   **Explained_by_Perceptions_of_corruption (Coefficient = 1.0669, p-value = 0.00865)**: This variable is significant, suggesting that lower perceptions of corruption are associated with higher happiness scores.

### **Comparative Analysis**:

-   **Which countries have the highest ‘Ladder score’?**

-   **How do factors like ‘Generosity’ and ‘Perceptions of corruption’ vary among the top 10 happiest countries compared to the bottom 10?**

    **Identify TOP 10 countries with highest ladder score:**

    ```{r}
    # Sort the data frame by Ladder_score in descending order
    sorted_data <- clean_Happy_data[order(-clean_Happy_data$Ladder_score), ]

    # Select the top 10 countries with the highest Ladder_score
    top_10_countries <- head(sorted_data, 10)

    # View the top 10 countries
    print("Top 10 happiest countries based on Ladder Score:")
    print(top_10_countries$Country_name)
    ```

    #### Analyze Generosity and Perceptions of corruption in Top 5 vs. Bottom 5 Countries

    ```{r}
    # Select the bottom 10 countries with the lowest Ladder_score
    bottom_10_countries <- tail(sorted_data, 10)



    # View the bottom 10 countries
    print("Bottom 10 countries based on Ladder Score:")
    print(bottom_10_countries$Country_name)

    # Compare Generosity and Perceptions of corruption
    top_10_comparison <- top_10_countries[, c("Country_name", "Explained_by_Generosity", "Explained_by_Perceptions_of_corruption", "Explained_by_Freedom_to_make_life_choices")]
    bottom_10_comparison <- bottom_10_countries[, c("Country_name", "Explained_by_Generosity", "Explained_by_Perceptions_of_corruption", "Explained_by_Freedom_to_make_life_choices")]

    # Print comparison for top 10
    print("Generosity and Perceptions of Corruption in the Top 10 happiest countries:")
    print(top_10_comparison)

    # Print comparison for bottom 5
    print("Generosity and Perceptions of Corruption in the Bottom 10 countries:")
    print(bottom_10_comparison)

    ```

    ```{r}
    # Combine top 5 and bottom 5 for comparison
    comparison_data <- rbind(top_10_countries, bottom_10_countries)
    comparison_data$Group <- rep(c("Top 10", "Bottom 10"), each = 10)
    ```

    ```{r}
    # Plot for Generosity
    ggplot(comparison_data, aes(x = reorder(Country_name, -Explained_by_Generosity), y = Explained_by_Generosity, fill = Group)) +
      geom_bar(stat = "identity", position = position_dodge()) +
      labs(title = "Generosity in Top 10 vs Bottom 10 Happiest Countries", x = "Country", y = "Generosity") +
      theme_minimal() +
       theme(axis.text.x = element_text(angle = 45, hjust = 1))+
      scale_fill_manual(values = c("Top 10" = "blue", "Bottom 10" = "red"))
    ```

    ```{r}
    # Plot for Perceptions of Corruption
    ggplot(comparison_data, aes(x = reorder(Country_name, -Explained_by_Perceptions_of_corruption), y = Explained_by_Perceptions_of_corruption, fill = Group)) +
      geom_bar(stat = "identity", position = position_dodge()) +
      labs(title = "Perceptions of Corruption in Top 10 vs Bottom 10", x = "Country", y = "Perceptions of Corruption") +
      theme_minimal() +
       theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      scale_fill_manual(values = c("Top 10" = "blue", "Bottom 10" = "red"))
    ```

    ```{r}
    # Plot for freedom to make life choices
    ggplot(comparison_data, aes(x = reorder(Country_name, -Explained_by_Freedom_to_make_life_choices), y = Explained_by_Freedom_to_make_life_choices, fill = Group)) +
      geom_bar(stat = "identity", position = position_dodge()) +
      labs(title = "Freedom make life choices in Top 10 vs Bottom 10", x = "Country", y = "Perceptions of Freedom") +
      theme_minimal() +
       theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      scale_fill_manual(values = c("Top 10" = "blue", "Bottom 10" = "red"))
    ```

    ### **Perceptions of Corruption Graph Analysis**

    1.  **Lower Corruption in Happier Countries**: The top 5 happiest countries (Denmark, Finland, Sweden, Israel, Iceland) have significantly lower perceptions of corruption compared to the bottom 5 (Afghanistan, Lesotho, Congo (Kinshasa), Sierra Leone, Lebanon). This indicates a trend where higher happiness levels correlate with lower perceived corruption.

    2.  **Uniformity Among Top Countries**: The top 5 countries show a relatively uniform low level of perceived corruption, suggesting a stable and trustworthy public sector which could be contributing to higher happiness scores.

-   **Variability Among Less Happy Countries**: The bottom 5 countries display a wider range of perceptions of corruption. This variability might reflect differing levels of governance and institutional effectiveness, which can affect societal happiness.

    ### **Generosity Graph Analysis**

    1.  **Comparative Generosity**: While both the top 5 and bottom 5 countries show variability in generosity, the top 5 happiest countries generally display higher levels of generosity compared to the bottom 5. This suggests that societal generosity could be associated with greater happiness.

    2.  **High Generosity Even in Less Happy Countries**: Notably, some of the less happy countries (e.g., Congo (Kinshasa) and Sierra Leone) still show significant levels of generosity. This indicates that despite economic or social challenges, cultural or societal norms in these countries may still promote generous behavior.

### **Distribution Analysis**:

-   **What is the distribution of the ‘Ladder score’ across all countries? Is it normally distributed?**

    ```{r}

    # Plot a histogram of Ladder_score
    hist(clean_Happy_data$Ladder_score, breaks = 20, main = "Histogram of Ladder Scores", xlab = "Ladder Score", col = "blue")

    # Q-Q plot for checking normality
    qqnorm(clean_Happy_data$Ladder_score, main = "Q-Q Plot of Ladder Scores")
    qqline(clean_Happy_data$Ladder_score, col = "red", lwd = 2)

    ```

    ### **Histogram Analysis:**

    -   **Shape of Distribution**: The histogram reveals a multi-modal distribution, indicating several peaks instead of one. The most pronounced peak is around a Ladder Score of 6, with smaller peaks around scores of 4 and 7.

    -   **Skewness**: The distribution appears to be slightly left-skewed as there are more scores on the higher end (right side) than on the lower end (left side). This skewness suggests that more countries are clustered around higher happiness scores.

    -   **Implication**: The presence of multiple peaks and slight skewness suggests that the Ladder Score distribution is not perfectly normal but instead shows variability in how happiness is distributed among countries.

    ### **Q-Q Plot Analysis:**

    -   **Linearity**: The Q-Q plot, which plots the quantiles of your Ladder Score data against the quantiles of a theoretical normal distribution, shows a mostly linear pattern, suggesting that the scores are relatively normally distributed.

    -   **Deviations**: However, note the slight deviations from the line at both the lower and higher ends of the distribution. These deviations are typical indicators of the heavy tails in the data — more extreme values (both low and high) than what a normal distribution would predict.

    -   **Normality Assessment**: While the central part of the data closely follows the theoretical line (indicating normality), the deviations at the tails suggest that the distribution is not perfectly normal. This is typical in real-world data, especially with complex phenomena like happiness.

    ### **Conclusion:**

    The **`Ladder Score`** distribution across countries shows characteristics of normality but with notable deviations, particularly in the form of slight skewness and heavier tails. This suggests that while many countries have happiness scores that approximate a normal distribution, there are countries at both ends of the spectrum that deviate from this trend. These findings highlight the complexity of factors influencing national happiness and suggest that multiple modes of happiness exist across different groups or types of countries.

-   **How does the distribution of ‘Healthy life expectancy’ compare to that of ‘Freedom to make life choices’?**

    ```{r}
    # Create histograms


    # Plot histograms for both variables
    par(mar=c(4, 4, 2, 2)) # Set up the plotting area to have 2 rows and 1 column

    # Histogram for Healthy life expectancy
    hist(clean_Happy_data$Explained_by_Healthy_life_expectancy, main="Histogram of Healthy Life Expectancy",
         xlab="Healthy Life Expectancy", col="blue", border="black")

    # Histogram for Freedom to make life choices
    hist(clean_Happy_data$Explained_by_Freedom_to_make_life_choices, main="Histogram of Freedom to Make Life Choices",
         xlab="Freedom to Make Life Choices", col="red", border="black")

    ```

    ```{r}
    # Reset the plotting area
    par(mfrow=c(1,1))

    # Plot density plots for both variables
    plot(density(clean_Happy_data$Explained_by_Healthy_life_expectancy), main="Density Comparison",
         xlab="Value", ylab="Density", col="blue", lwd=2, xlim=c(min(c(clean_Happy_data$Explained_by_Healthy_life_expectancy, clean_Happy_data$Explained_by_Freedom_to_make_life_choices)),
         max(c(clean_Happy_data$Explained_by_Healthy_life_expectancy, clean_Happy_data$Explained_by_Freedom_to_make_life_choices))))
    lines(density(clean_Happy_data$Explained_by_Freedom_to_make_life_choices), col="red", lwd=2)
    legend("topleft", legend=c("Healthy Life Expectancy", "Freedom to Make Life Choices"),
           col=c("blue", "red"), lty=1, lwd=2)
    ```

### **Histogram Analysis**

1.  **Healthy Life Expectancy**:

    -   The histogram for **`Healthy Life Expectancy`** is slightly right-skewed, with a major peak around 0.6. This suggests that most countries have a higher value for healthy life expectancy.

    -   The distribution is not uniform and shows concentration in the mid to higher range, indicating variability in how different countries achieve healthy life expectancies.

2.  **Freedom to Make Life Choices**:

    -   In contrast, the histogram for **`Freedom to Make Life Choices`** shows a roughly symmetric distribution with a strong central peak around 0.6 to 0.8. This indicates a clustering of countries where citizens perceive they have substantial freedom in making life choices.

    -   The presence of smaller bars at both the lower and higher ends suggests that while many countries have high freedom, a few countries either have very high or very low perceived freedoms.

### **Density Plot Analysis**

-   **Overlap and Distinction**:

    -   The density plot provides a clearer visualization of how each distribution spreads out and peaks. Both variables share a similar range, but their distributions tell different stories.

    -   **Healthy Life Expectancy** peaks earlier and has a longer tail towards higher values, consistent with the right skew seen in the histogram. This could reflect disparities in health services and outcomes among countries.

    -   **Freedom to Make Life Choices** shows a more bell-shaped curve, peaking later than Healthy Life Expectancy and tapering off more symmetrically. This distribution is closer to normal, suggesting that perceptions of freedom are more uniformly distributed among the countries surveyed.

### **Interpretation and Implications**

-   The comparison highlights that while there is a general global trend towards higher healthy life expectancies and freedom to make life choices, the variation in healthy life expectancy is more pronounced. This might be due to differing healthcare policies, economic conditions, and public health realities.

-   The more uniform distribution of freedom to make life choices might suggest a global convergence towards recognizing and ensuring these freedoms, although extremes still exist.

## Data source {.unnumbered}

Where does the data come from, who owns the data? Where are all the scripts that you need to clean the data?

Data Source: <https://worldhappiness.report/data/>

The data used is the 2024 data for Figure 2.1. The World Happiness Report, from which this data originates is, is a partnership between Gallup, the Oxford Wellbeing Research Centre, the UN Sustainable Development Solutions Network, and the WHR’s Editorial Board. It is also important to note that the report is produced under the editorial control of the WHR Editorial Board.

## Conclusion {.unnumbered}

While working on this project it was realized that this report only includes a small number of variables that interpret happiness and it is a generalization across each country citizen's. Therefore, things that might make one individual happy might not make another. With this in mind it is necessary to consider other relevant factors to you before making an informed decision on where you might wish to move, such as racism, homophobia climate, visas, etc. before deciding to moving anywhere.

## References {.unnumbered}

-   Linear Regression: <https://r-statistics.co/Linear-Regression.html>

-   Pearson Correlation in R: <https://www.geeksforgeeks.org/pearson-correlation-testing-in-r-programming/>

-   Q-Q plot:<https://whitlockschluter3e.zoology.ubc.ca/Tutorials%20using%20R/R_tutorial_Normal_and_sample_means.html>
